Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-05-12 closed_loop_planning.tex
[MGa] mutually exclusive throttle and braking command
how to handle

Given the nominal trajectory $(\hbmu_k, \hbu_k)$ over $k=0,\ldots, N$, planned at step (i), and introduced the deviation variables $\bbmu_k=\bmu_k - \hbmu_k$ and $\bbu = \bu_k - \hbu_k$, we design a linear state-feedback law $\bbu_k= -\hbK_k \bbmu_k$, or equivalently $\bu_k = \hbu_k - \hbK_k (\bmu_k - \hbmu_k)$, to mitigate disturbances while tracking the nominal trajectory.

@@

>>>>

Given the nominal trajectory $(\hbmu_k, \hbu_k)$ over $k=0,\ldots, N$, planned at step (i), and introduced the deviation variables $\bbmu_k=\bmu_k - \hbmu_k$ and $\bbu_k = \bu_k - \hbu_k$, we design a linear state-feedback law $\bbu_k= -\hbK_k \bbmu_k$, or equivalently $\bu_k = \hbu_k - \hbK_k (\bmu_k - \hbmu_k)$, to mitigate disturbances while tracking the nominal trajectory.

@@

\subsection{Handling mutually exclusive throttle and braking commands}
In our formulation, the state vector is defined as $\bx=\big[u\;v\;r\;x_{G}\;y_{G}\;\psi \big]^T\in\bbR^6$, while the control input vector is $\bu=\big[F_x^a\;F_x^b\;\de \big]^T\in\bbR^3$. The longitudinal acceleration and braking forces, $F_x^a \geq 0$ and $F_x^b \leq 0$, respectively, are not independent but are subject to the \emph{complementarity constraint} $F_x^a F_x^b = 0$, which enforces mutual exclusivity between throttle and braking at any instant.

For convenience, we denote $u_1 = F_x^a \geq 0$ and $u_2 = F_x^b \leq 0$, so that the complementarity condition becomes $u_1 u_2 = 0$. Beyond the modeling constraint itself, this condition also imposes structural implications on admissible control variations, particularly relevant in the closed-loop formulation. 
Specifically, in a nominal \emph{acceleration} phase ($\hat{u}_1 > 0, \hat{u}_2 = 0$), both \(\delta u_1 \lesseqgtr 0\) are permitted, but only variations $\delta u_2 < 0$ are physically meaningful -- since a positive brake increment ($\delta u_2 > 0$) would unphysically contribute to propulsion. Conversely, in a \emph{braking} phase ($\hat{u}_2 < 0$, $\hat{u}_1 = 0$), $\delta u_2 \lesseqgtr 0$ are admissible, but only $\delta u_1 > 0$ should be allowed, as negative throttle increments ($\delta u_1 < 0$) would again introduce non-physical behavior.

To handle this, we introduce a \emph{minimal control vector} \(\tilde{\bu} = [F_x\;\; \delta]^T \in \mathbb{R}^2\), where \(F_x\) represents the net longitudinal force. The full control vector $\bu$ is then recovered via a state-dependent linear transformation $
\bu = \bH(\tilde{\bu}) \tilde{\bu}$ with
\begin{equation} \label{eq:minimalcontrol}
\bH(\tilde{\bu}) =
\begin{bmatrix}
\Ipos{F_x} & 0 \\
\Ineg{F_x} & 0 \\
0 & 1
\end{bmatrix}, \quad
\Ipos{x} :=
\begin{cases}
1, & \text{if } x \geq 0 \\
0, & \text{if } x < 0
\end{cases}, \quad
\Ineg{x} :=
\begin{cases}
0, & \text{if } x \geq 0 \\
1, & \text{if } x < 0
\end{cases}.
\end{equation}

In this formulation, \(\Ipos{x}\) and \(\Ineg{x}\) denote the \emph{positive-} and \emph{negative-part indicator functions}, respectively. Importantly, control variations also respect this mapping: $\delta\bu = \bH(\tilde{\bu}) \delta\tilde{\bu}$. Consequently, the control input matrix can be expressed as \(\bB \bH = \tilde{\bB}\), and the state-feedback gain as \(\bH \tilde{\bK} = \bK\), such that the resulting feedback law satisfies \(\bB \bK \delta\bx = \tilde{\bB} \tilde{\bK} \delta\bx\) and $\de \tbu = \tbK \de \bx$.

To correctly enforce the mutual exclusivity within the LQR synthesis via Riccati recursion, the state-feedback gain must be computed with respect to the minimal control input. Specifically, the control input matrix $\tbB$ is employed and the gain $\tilde{\bK} \in \mathbb{R}^{2 \times 6}$ is sought for which operates on the independent variation \(\delta\tilde{\bu}\). This is structured as
\[
\tilde{\bK} = \begin{bmatrix} \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T, \quad \tilde{\bK}_F, \tilde{\bK}_\delta \in \mathbb{R}^{1 \times 6}.
\]
The extended gain $\bK$ is automatically recovered via $\bK = \bH \tbK$ and, depending on the operating mode, it will result in the following expressions:
\begin{itemize}
  \item during \emph{acceleration}: \(\bK = \begin{bmatrix} \tilde{\bK}_F^T & \mathbf{0}^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\),
  \item during \emph{braking}: \(\bK = \begin{bmatrix} \mathbf{0}^T & \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\).
\end{itemize}
This construction ensures consistency with the complementarity constraint, while maintaining a tractable structure for controller synthesis.

A similar consideration applies when solving problem~\eqref{eq:ROCP} in Sec.~\ref{sec:CLrobustified}, where the complementarity structure must again be taken into account. In this case, it is sufficient to treat the elements of $\tbK_k$ as the minimal decision variables and reconstruct the full gain matrix $\bK_k$ via the mapping $\bK_k = \bH_k^{\text{S}} \tbK_k$, where
\begin{equation} \label{eq:minimalcontrolopt}
\bH^{\text{S}}(\tilde{\bu}) =
\begin{bmatrix}
\SIpos{u_1+u_2} & 0 \\
\SIneg{u_1+u_2} & 0 \\
0 & 1
\end{bmatrix}. 
\end{equation}
To ensure differentiability and numerical smoothness in the optimization problem~\eqref{eq:ROCP}, we employ smoothed approximations of the indicator functions $\Ipos{x}$ and $\Ineg{x}$, denoted as $\SIpos{x}$ and $\SIneg{x}$ in~\eqref{eq:minimalcontrolopt}, respectively. These are defined as:
\begin{equation} \label{eq:smoothedindicator}
\SIpos{x} :=
\frac{1}{2} \left(1 + \tanh(\chi x)\right), \quad \SIneg{x} := \frac{1}{2}\left(1 - \tanh(\chi x) \right),
\end{equation}
where the parameter $\chi>0$ controls the sharpness of the transition. Smaller values of $\chi$ yield smoother, more relaxed approximations, while larger values make the transition steeper and the function increasingly resemble a hard switching indicator.

  %However, longitudinal acceleration and braking forces $F_x^a \geq 0$ and $F_x^b \leq 0$ are not independent, but are subjected to the complementarity constraint $F_x^a F_x^b=0$, meaning that are mutually exclusive. Introduced for simplicity $u_1=F_x^a\geq 0$ and $u_2 = F_x^b\leq 0$, at any instant in time it must hold that $u_1 u_2 = 0$. What is important is also that such constraints have also an implication on the values that control variations associated to closed-loop types of correction can assume. More in details, in a condition of acceleration (reference acceleration $\hat{u}_1>0$ and $\hat{u}_2 = 0$) $\de u_1\lesseqgtr 0$ but only $\de u_2 <0$ is allowed, otherwise a brake variation $\de u_2 > 0$ would unphysically help the vehicle to accelerate. Similarly, in a braking condition (reference braking $\hat{u}_2 < 0$ and $\hat{u}_1 = 0$) $\de u_2 \lesseqgtr 0$ but only $\de u_1 > 0$, otherwise a variation of acceleration $\de u_1 < 0$ would unphysically help the vehicle to brake. This brings about the necessity to properly \emph{under-actuate} the system in order account for this behavior. This can be accomplished by introducing the minimal control $\tilde{\bu}=[F_x, \de]^T$ such that full control with mutually exclusive acceleration/braking can be recovered as $\bu = \bH(\tbu) \tbu$ where
%\begin{align}\label{eq:minimalcontrol}
%\bH(\tbu) =
%\begin{bmatrix}
%\Ipos{F_x} & 0\\
%\Ineg{F_x} & 0\\
%       0 & 1
%\end{bmatrix}, \, \text{with} \quad
%\Ipos{x} :=
%\begin{cases}
%1, & \text{if } x \geq 0 \\
%0, & \text{if } x < 0
%\end{cases}, \quad \Ineg{x} :=
%\begin{cases}
%0, & \text{if } x \geq 0 \\
%1, & \text{if } x < 0
%\end{cases}.
%\end{align}
%In~\eqref{eq:minimalcontrol} we introduced the positive- and negative-part indicator functions $\Ipos{x}$ and $\Ineg{x}$, respectively. It is worth observing that the extended control variation $\de \bu = \bH(\tbu) \de \tbu$.
% These following positions can be made $\bB \bH = \tbB$, $\bH \tbK = \bK$, such that $\bB \bK \de \bx = \tbB \tbK \de \bx$.
% It is worth observing that in order to account for the mutually exclusive condition in the synthesis of the LQR controller via the Riccati recursion, the state feedback gain matrix $\tbK\in\bbR^{2\times 6}$ must be employed. This allows us to follow the classical recursion by operating on the fully independent $\de \tbu$ and it will result in the calculation of the minimal state-feedback gain matrix $\tbK=\big[ \tbK_{F}^T\, \tbK_{\de}^T\big]^T$ with $\tbK_{F}\in\bbR^{1\times 6}$ and $\tbK_{\de}\in\bbR^{1\times 6}$. Then, in the extended controls it will be possible to recover $\bK$ such that during acceleration $\bK=\big[ \tbK_{F}^T\, \bzero^T\, \tbK_{\de}^T\big]^T$  and, during braking, $\bK=\big[ \bzero^T\, \tbK_{F}^T\, \tbK_{\de}^T\big]^T$.
